---
2222---
title: "Almond Yield Anomalies"
author: "Jackson Coldiron and Kaitlin Castor"
editor: visual
format: 
  html:
    code-fold: true
    embed-resources: true
    theme: lumen
    toc: true
execute:
  freeze: auto 
  warning: false
  message: false
---

## Introduction
This assignment builds on our almond yield anomaly model by incorporating an economic dimension—profit. Using yield anomalies calculated from climate data, we developed a profit model that estimates annual almond profits based on variable production costs and market prices. We then performed a simple sensitivity analysis to explore how uncertainty in key parameters—specifically cost per acre and price per ton—affects estimated profit across years. This analysis allows us to assess the robustness of almond profitability under different economic scenarios and visualize how sensitive outcomes are to these inputs.

## Data
#### Data Description
The climate dataset (clim.txt) contains daily weather observations including:

day: Day of the month

month: Calendar month

year: Calendar year

wy: Water year

tmax_c: Daily maximum temperature (°C)

tmin_c: Daily minimum temperature (°C)

precip: Daily precipitation (mm)

Other parameters: - location: California - time frame: 1989-2010 - yield model adjusted R\*\*2 = 0.88 - yield model developed from monthly averages aggregated to the state scale from 1980 - 2003 records. This excludes factors such as extreme heat events or spatial variations.

These inputs are used to compute annual almond yield anomalies. The expected outputs include the maximum, minimum, and mean yield anomalies across all years in the dataset. According to the assignment benchmark, the modeled yield anomalies should approximately range between –0.027 to 1920 tons/acre, with a mean around 182 tons/acre.

#### Data Processing

```{r}
# Import libraries
library(tidyverse)
library(here)
library(dplyr)
library(purrr)
library(tibble)
library(knitr)
library(kableExtra)

# Define baseline yield
baseline <- 5

# Import data and function
climate_data_raw <- read.table(here("data", "clim.txt"), header = TRUE, sep = "", quote = "\"")

source(here("R", "almond_yield_function.R"))
source(here("R", "almond_profit_function.R"))

# Summarize inputs by year for model
yearly_anomalies <- climate_data_raw |>
  group_by(year) |>
  summarise(
    Tn = mean(tmin_c[month == 2], na.rm = TRUE),
    P = sum(precip[month == 1], na.rm = TRUE)) |>
    drop_na() |>
  mutate(yield_anomaly = almond_yield(Tn, P))

```

## Almond Profit

# Apply the almond wrapper function to each year

```{r}
# Bring in the wrapper function
source(here("R", "almond_wrapper_function.R"))

# Apply wrapper function on each year
yearly_profits <- yearly_anomalies |>
  mutate(
    profit = pmap_dbl(
      list(Tn, P),
      ~ almond_profit_wrapper(Tn, P)
    )
  )
```

## Sensitivity Analysis

#### Cost per ton Sensitivity - Kaitlin

```{r}
# generate samples for both parameters
nsamples <- 300
deviation <- 0.15
base_cost <- 15000

set.seed(42)  # for reproducibility

costs_per_acre <- runif(
  min = base_cost - deviation * base_cost,
  max = base_cost + deviation * base_cost, n = nsamples
)

price_per_ton <- rnorm(mean = 4000, sd = 0.1, n = nsamples)

parms <- cbind.data.frame(costs_per_acre, price_per_ton)

# create dataframe with all combinations of parameters
sensitivity_df <- crossing(
  yearly_anomalies,
  parms
)

# run the profit function across the new dataframe
sensitivity_profit <- sensitivity_df |>
  mutate(
    profit = pmap_dbl(
      list(price_per_ton, costs_per_acre, yield_anomaly),
      ~ almond_profit(
          price_per_ton = price_per_ton,
          costs_per_acre = costs_per_acre,
          baseline_yield = baseline,
          yield_anomaly = yield_anomaly
        )
    )
  )

```

#### Figure 1: Comparison of parameters sensitivity on profit

```{r}
# plot - pick on of the 2 parameter as a color
p1 <- ggplot(mean_profit, aes(costs_per_acre, mean, col = price_per_ton)) +
  geom_point(cex = 2) +
  labs(y = "Mean Annual Profit $", x = "Cost per Acre ($/acre)  \n above which energy production is more efficient")
p2 <- ggplot(mean_profit, aes(price_per_ton, mean, col = costs_per_acre)) +
  geom_point(cex = 2) +
  labs(y = "Mean Annual Profit $", x = "Price per tone $")
ggarrange(p1, p2)

# what do we learn from this

# extract annual #I did not touch this yet
tmp <- map_df(results, `[`, c("annual"))
annual_elect <- as.data.frame(tmp$annual$year)
colnames(annual_elect) <- "year"
annual_elect$elect <- tmp$annual$elect
```

#### Figure 2: Overall Variation of Yield Anomaly

## Discussion
